---
title: Age-Sex Pyramids with Python and R!
author: Carlos
date: '2018-07-29'
slug: age-sex-pyramids-with-python-and-r
categories:
  - Census
  - Analysis
tags:
  - Census
  - Python
  - R
subtitle: ''
---

The _reticulate_ package was made available earlier this year. This package allows us to use R and Python within RStudio! 

In this post, I provide an example of using Python to extract data from the US Census API to construct a population pyramid for Bexar County. Population pyramids are a good way for analyzing the age-sex composition of a population, which can give us an idea if a population is younger/older. I understand that one can make a call using to the Census API with R, but I prefer using Python for certain tasks :).

First, let's load the reticulate package 

```{r}

library(reticulate)

```

Now, I invoke python within this markdown file and run commands to extract data from the Census API. Basically, I will get all of the population estimates by age and sex.

```{python}

import pandas as pd

# List of variables
var = 'B01001_'

var_list = []

for i in range(1,50):
    tmp_var = var + str(i).zfill(3) + 'E'
    
    var_list.append(tmp_var)

var_list_s = str(var_list).strip('[]')
var_list_p = ','.join(var_list)

```

I create a variable to get the age-sex population estimates data for Bexar County Texas and will store as a pandas dataframe.

```{python}

# Census API call
url_link = 'https://api.census.gov/data/2016/acs/acs5?get=NAME,' + var_list_p + '&for=county:029&in=state:48&key=dbcf37f4c47be0a2a720ec01f7c32e6792846be0'
       
# Get block group info
county_data = pd.read_json(url_link)
county_data = pd.DataFrame(county_data[1:])

```

Now that we have our pandas dataframe, let's take a look at the data

```{python}

print(county_data.head())

```

We see that we need to rename the columns.

```{python}

column_names = ['NAME']
column_names.extend(var_list)
geo_names = ['STATE','COUNTY']
column_names.extend(geo_names)

county_data.columns = column_names

print(county_data.head())

# Output to .csv file

county_data.to_csv('../..//content/data/bexar_county_age_sex_pop.csv')

```

Great! 

Let's use R to create our population pyramids. 

Load the _dplyr_ and _ggplot2_ packages

```{r, echo=F}

suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(readr))

```

Then, we read in the .csv file we created with Python and do some data manipulation by renaming columns, reshaping the dataset, and plotting

```{r}

county_data <- suppressMessages(read_csv('../..//content/data/bexar_county_age_sex_pop.csv'))

# Create age group list
age_gp <- c('< 5', '5-9','10-14','15-19','20-24','25-29',
            '30-34','35-39','40-44','45-49','50-54','55-59',
            '60-64','65-69','70-74','75-79','80-84','85+')

county_data <- county_data %>%
  mutate(pop_15to19_m = B01001_006E + B01001_007E,
         pop_20to24_m = B01001_008E + B01001_009E + B01001_010E,
         pop_60to64_m = B01001_018E + B01001_019E,
         pop_65to69_m = B01001_020E + B01001_021E,
         
         pop_15to19_f = B01001_030E + B01001_031E,
         pop_20to24_f = B01001_032E + B01001_033E + B01001_034E,
         pop_60to64_f = B01001_042E + B01001_043E,
         pop_65to69_f = B01001_044E + B01001_045E,
         FIPS = paste(STATE, COUNTY, sep='')) %>%
  
  select(FIPS, B01001_001E, B01001_003E:B01001_005E, 
         pop_15to19_m, pop_20to24_m, B01001_011E:B01001_017E,
         pop_60to64_m, pop_65to69_m, B01001_022E:B01001_025E,
         B01001_027E:B01001_029E, 
         pop_15to19_f, pop_20to24_f, B01001_035E:B01001_041E,
         pop_60to64_f, pop_65to69_f, B01001_046E:B01001_049E) %>%
  
  tidyr::gather(key = Variable, value = Pop_Count, -FIPS, -B01001_001E) %>%
  
  mutate(var_order = rep(1:18, each=2)) %>%
  
  arrange(FIPS, var_order) %>%
  
  mutate(Age_Group = factor(rep(1:18, 2), 
                            labels=age_gp),
         Sex = factor(rep(rep(1:2, each=18)),
                      labels=c('Male','Female')),
         pct_pop = Pop_Count / B01001_001E,
         pct_pop = ifelse(Sex == 'Male', pct_pop*-1,
                          pct_pop))

head(county_data)

```

Plot!

```{r}
ggplot(county_data,
       aes(x = Age_Group, y = pct_pop, fill = Sex)) + 
  geom_bar(subset = (county_data$Sex == "Female"), stat = "identity") + 
  geom_bar(subset = (county_data$Sex == "Male"), stat = "identity") + 
  coord_flip() + 
  scale_fill_brewer(palette = "Set1") + 
  theme_bw() +
  ylab('Percent') +
  xlab('Age Group')

```